Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-05-31 elsevier_MANUSCRIPT.qmd
new draft

We tested multiple model configurations, including versions with the full predictor set across the state, separate models for physiographic regions with full predictor set, and reduced subsets across the state, to evaluate the impact of different predictor combinations on model performance and to ensure robustness of the final selected model. Dimensionality reduction was performed to improve interpretability and reduce computational burden. We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors. SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions. The ten most important features, based on SHAP value, were retained in a reduced model. The reduced model showed improved predictive accuracy and efficiency over the full model and regional models. Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features outperformed alternatives. The final model was applied to all HUC8 basins across Arizona using the same predictor layers to estimate long-term BFI values.

![SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values alt="SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group."}

The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] . In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds being rotated until all of the data had been used for validation in separate models. Root mean squared error (RMSE) was used as the primary performance metric. Model performance on unseen data was strong, with an overall R² of 0.764. Additional model validation results are presented in section 3.2.1.

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold alt="Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds." fig-align="center" width="100%"}

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

>>>>

We tested multiple model configurations, including versions with the full predictor set across the state, separate models for physiographic regions with full predictor set, and reduced subsets across the state, to evaluate the impact of different predictor combinations on model performance and to ensure robustness of the final selected model. Dimensionality reduction was performed to improve interpretability and reduce computational burden. We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors. SHAP was chosen for its interpretability and ability to account for feature interactions [@SHAP_values]. The values quantify both the magnitude and direction of each feature’s contribution to predictions. The ten most important features, based on SHAP value, were retained in a reduced model. The reduced model showed improved predictive accuracy and efficiency over the full model and regional models. Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features outperformed alternatives. The final model was applied to all HUC8 basins across Arizona using the same predictor layers to estimate long-term BFI values.

![SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] . In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds being rotated until all of the data had been used for validation in separate models. Root mean squared error (RMSE) was used as the primary performance metric. Model performance on unseen data was strong, with an overall R² of 0.764.

To contextualize model accuracy, we compared XGBoost to two baseline methods: a simple linear regression and an inverse distance weighting (IDW) interpolation. XGBoost reduced RMSE by 43% relative to the linear model and 30% relative to IDW, while improving NSE from 0.25 (linear) and 0.45 (IDW) to 0.75, indicating substantially stronger predictive accuracy and spatial generalization.

![Schematic of \$k\$-fold cross validation. The dataset is randomly divided into \$k\$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all \$k\$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|-------------------------|-------|-------|-------|-------|-------|-------|-------|
| Full Dataset | 7724 | 0.764 | 0.015 | 0.122 | 0.081 | 0.752 | -5.6 |
| linear model | 7724 | 0.248 | 0.046 | 0.214 | 0.173 | 0.247 | \~ 0 |
| Inverse Distance Weighted (LOO CV) | 7724 | 0.449 | 0.029 | 0.170 | 0.131 | 0.448 | 1.86 |
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

The reduced predictor set was selected to maximize interpretability and minimize overfitting. However, incorporating additional variables could improve model performance, especially in capturing regionally complex baseflow dynamics. Subsurface hydrologic characteristics, such as aquifer properties, groundwater depth, and spring density, are known to influence baseflow but were excluded due to limited availability across the study domain. Temporal predictors, including snowpack duration, antecedent soil moisture, and wetness indices, could improve annual BFI predictions by capturing lagged hydrologic responses to climate variability. Additionally, incorporating human influences (for example, groundwater withdrawals, land-use change, or reservoir operations) would be important in regions with modified hydrologic regimes. Future efforts may benefit from integrating these datasets where available, although such improvements would require region-specific data processing and validation strategies.
